Prosper Loan Data Analysis by John Lauritzen
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(gridExtra)
library(RColorBrewer)
library(dplyr)
library(reshape2)
library(lubridate)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
pl <- read.csv('prosperLoanData.csv')
# Formatting dates to be recognized as dates
pl$ListingCreationDate <- strptime(pl$ListingCreationDate,
                                   format = "%Y-%m-%d %H:%M:%S")
pl$ClosedDate <- strptime(pl$ClosedDate, format = "%Y-%m-%d %H:%M:%S")
pl$DateCreditPulled <- strptime(pl$DateCreditPulled, 
                                format = "%Y-%m-%d %H:%M:%S")
pl$FirstRecordedCreditLine <- strptime(pl$FirstRecordedCreditLine, 
                                       format = "%Y-%m-%d %H:%M:%S")
pl$LoanOriginationDate <- strptime(pl$LoanOriginationDate, 
                                   format = "%Y-%m-%d %H:%M:%S")
# Factoring variables that have an inherent order
pl$CreditGrade <- factor(pl$CreditGrade)
pl$ProsperRating..numeric. <- factor(pl$ProsperRating..numeric.)
pl$ProsperScore <- factor(pl$ProsperScore)
pl$IncomeRange <- factor(pl$IncomeRange)
pl$Term <- factor(pl$Term)
# Specifying ordered factoring for variables without inherent order
pl$IncomeRange <- ordered(pl$IncomeRange, levels = c("Not employed", "$0",
                                                     "$1-24,999",
                                                     "$25,000-49,999",
                                                     "$50,000-74,999",
                                                     "$75,000-99,999",
                                                     "$100,000+",
                                                     "Not displayed"))
pl$ProsperRating..Alpha. <- ordered(pl$ProsperRating..Alpha.,
                                    levels = c("", "HR", "E", "D", "C", "B",
                                               "A", "AA"))
pl$CreditGrade <- ordered(pl$CreditGrade, levels = c("", "HR", "E", "D", "C",
                                                     "B", "A", "AA"))
# Creating new variables for ease of use in plotting
pl$AverageCreditScore <- (pl$CreditScoreRangeLower +
                                          pl$CreditScoreRangeUpper)/2
pl$LoanOriginationYear <- factor(year(pl$LoanOriginationDate))
```

This report explores a dataset containing loans and customer information for 
approximately 113,000 loans from Prosper, a peer-to-peer lending marketplace.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot}
qplot(data = pl, x = LoanOriginationYear, binwidth = 1)
```

First, looking at the years of the originated dates, we can see that the loans 
contained in the data range from 2005 to 2014.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot1}
# Create standard to adjust x label to prevent overlap  
x45 <- theme(axis.text.x = element_text(angle = 45, hjust = 1))

qplot(data = pl, x = LoanOriginalAmount, binwidth = 1000) +
  scale_x_continuous(breaks = seq(0, 40000, 2000), limits = c(0, 36000)) +
  x45
```

Looking at the spread of the original loan amounts we see several distinct 
spikes in loan amount around 4,000, 10,000, 15,000, and smaller spikes at 
20,000 and 25,000. We also see most of the loan amounts clustered under 10,000, 
which indicates to me these are more along the lines of personal loans and not 
other types of loans like mortgages. So what are the people like who are 
getting these loans?

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot2}
# Plot using subset to exclude not displayed income ranges
qplot(data = subset(pl, IncomeRange != "Not displayed"), x = IncomeRange) +
  x45
```

Looking at the income range of the customers, we can see the majority of them 
earn between 25,000 and 74,999.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot3}
qplot(data = pl, x = AverageCreditScore, binwidth = 20) +
  scale_x_continuous(breaks = seq(0, 900, 50))
```

Looking at the average credit score, calculated by averaging the lower and upper 
range of the customer's scores, we see that most of these loans were made to 
people in the range of 625 to 750.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Table}
table(pl$Occupation)
```

There is a wide variety of professions listed for these people, though most 
fall under the generic terms of Professional or Other. So how have these loans 
worked out for the investors who provide the funds?

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, Univariate_Plot4}
qplot(data = pl, x = LoanStatus) +
  x45
```

With most being Completed or Current things seem to be going pretty well.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot5}
qplot(data = pl, x = EstimatedReturn, binwidth = 0.025) +
  scale_x_continuous(breaks = seq(-0.1, 0.25, 0.05), limits = c(-0.1, 0.25))

summary(pl$EstimatedReturn)
```

One of the measures of benefit to the investors presented is the 
Estimated Return, which is the difference between the Estimated Effective Yield 
(Interest - Service Fee - charge-offs + late fees) and the Estimated Loss Rate 
(principal loss on charge-off). I'm surprised to find that there are loans with 
an Estimated Return less than 0.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Table1}
summary(pl$Investors)

qplot(data = pl, x = Investors, binwidth = 10)
```

Taking a look at the summary of investors, the count of how many people 
invested to provide the funds for the loan, we see a median of 44 which seems 
to be more accurate than the mean due to the long tail that can be seen when 
charting the data.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot6}
qplot(data = pl, x = MonthlyLoanPayment, binwidth = 50) +
  scale_x_continuous(breaks = seq(0, 1500, 100), limits = c(0, 1500))
```

The most common monthly payment is around 150. I find it interesting that this 
plot follows a similar shape to the previous loan amount, but without the 
distinct spikes at higher values.

# Univariate Analysis

### What is the structure of your dataset?
There are a total of 113,937 loans with 81 variables. Some of these are specific
for data management, such as ListingKey, but the majority of the variables 
contain relevant data for analysis. Of the data, the variables 
ProsperRating..Alpha. and CreditGrade are ordered factor variables with the 
following levels going from worst to best:

1. Blank
2. HR
3. E
4. D
5. C
6. B
7. A
8. AA

These two variables contain the same values, but CreditGrade is only relevant to
loans prior to 2009, and ProsperRating..Alpha. applies to loans after July 2009.
This indicates there was a change in how they calculated ratings in 2009, but 
they kept the same rating values.

Other Observations:

* There are distinct clusters of loans around 4,000, 10,000, 15,000, 20,000, 
and 25,000 dollars
* Most borrowers report yearly incomes between 25,000 and 74,999
* Most borrowers have an average credit score between 625 and 750
* The average estimated return for investors is about 9%
* The median number of investors per loan is 44
* Monthly payment amount follows a similar pattern to loan amount, but without 
the distinct clusters other than the mode

### What is/are the main feature(s) of interest in your dataset?
I think the main features of the data depend on which side you are looking at 
it from. As a borrower, the main features would be LoanOriginalAmount and 
MonthlyLoanPayment to determine if a loan through Prosper would be competitive 
to other options. As an investor the main features would be EstimatedReturn 
and LoanOriginalAmount.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
For borrowers, items such as the AverageCreditScore, DebtToIncomeRatio, 
and ProsperScore probably contribute heavily to the terms offered. For 
investors the items such as the ProsperScore, AverageCreditScore, and 
DelinquenciesLast7Years are most likely good indicators of good investments.

### Did you create any new variables from existing variables in the dataset?
Yes, I created an AverageCreditScore variable, derived by averaging the 
CreditScoreRangeLower and CreditScoreRangeUpper to create a single value for use
 in graphing and calculations. I also created LoanOriginationYear, derived from 
LoanOriginationDate, in order to easily graph off of the year the loan 
originated.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The results of the LoanOriginalAmount had an unusual distribution with several 
large clusters of values around 4,000, 10,000, 15,000, 20,000, and 25,000 
dollars. The only data I adjusted was excluding "Not displayed" from the 
IncomeRange graph.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot}
ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount), data = pl) +
  geom_point(alpha = 1/20)

summary(lm(MonthlyLoanPayment ~ LoanOriginalAmount, data=pl))
```

There seem to be 4 main patterns of monthly payment that emerge, which I suspect
is going to be related to the credit score of the borrowers.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot1}
ggplot(aes(x = MonthlyLoanPayment, y = AverageCreditScore), data = pl) +
  geom_jitter(alpha = 1/20) +
  scale_y_continuous(breaks = seq(425, 900, 25), limits = c(400, 900))
# Limit values to better focus the plot

summary(lm(MonthlyLoanPayment ~ AverageCreditScore, data=pl))
```

We see a somewhat similar pattern emerge with the majority of the loan payments 
clustered pretty close together up to a little beyond 500 a month, and then to 
decreasing groupings beyond that. Credit score does seem to have an impact as 
there are distinct steps up at specific credit score averages.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot2}
ggplot(aes(x = LoanOriginalAmount, y = AverageCreditScore), data = pl) +
  geom_jitter(alpha = 1/20) +
  scale_y_continuous(breaks = seq(425, 900, 25), limits = c(400, 900))
# Follow the same limits as previous plot for consistency

summary(lm(LoanOriginalAmount ~ AverageCreditScore, data=pl))
```

The groupings here are a bit less pronounced, but that matches what we saw 
previously where the main gap between the two lines happened after the 25,000 
loan amount. So how does the Prosper Score compare to the average credit score?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot3}
# Subset of data to exclude records without Prosper score
ggplot(aes(x = ProsperScore, y = AverageCreditScore),
       data = subset(pl, ProsperScore != "NA")) +
  geom_boxplot()

summary(lm(AverageCreditScore ~ ProsperScore, data=pl))
```

For the most part this matches what I expected, with average scores increasing
as Prosper Score increases, but I'm surprised by the dip at 11 which is the 
highest score. Perhaps previous delinquencies have an impact on the score.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot4}
ggplot(aes(x = ProsperScore, y = DelinquenciesLast7Years), 
       data = subset(pl, ProsperScore != "NA")) +
  geom_jitter(alpha = 1/50)

ggplot(aes(x = ProsperScore, y = DelinquenciesLast7Years), 
       data = subset(pl, ProsperScore != "NA")) +
  geom_boxplot()
```

We do see a distinctly smaller amount of delinquencies among those with a score 
of 11. Another item that may have a large impact would be the debt to income
ratio.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot5}
ggplot(aes(x = ProsperScore, y = DebtToIncomeRatio), 
       data = subset(pl, ProsperScore != "NA")) +
  geom_boxplot() +
  scale_y_continuous()

ggplot(aes(x = ProsperScore, y = DebtToIncomeRatio), 
       data = subset(pl, ProsperScore != "NA")) +
  geom_boxplot() +
# Limit y scale to better see variation in values
  scale_y_continuous(breaks = seq(0, 1, 0.05), limits = c(0, 1))

summary(lm(DebtToIncomeRatio ~ ProsperScore, data=pl))
```

Again, we see a distinct decrease in debt to income ratio as the Prosper Score 
increases though again we see a slight variation when it comes to the 11 score. 
One question I have is if there is a reliable relationship between the debt to 
income ratio and the amount of delinquencies?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot6}
ggplot(aes(x = DebtToIncomeRatio, y = DelinquenciesLast7Years), 
       data = pl) +
  geom_jitter(alpha = 1/20) +
# Follow limitation from previous plot for consistency
  scale_x_continuous(breaks = seq(0, 1.0, 0.05), limits = c(0, 1.0))

summary(lm(DebtToIncomeRatio ~ DelinquenciesLast7Years, data=pl))
```

This honestly surprises me. I would expect a higher dept to income ratio to have
 a higher number of delinquencies. While we do slightly see that trend at first 
it reverses at around 15%.

Clearly a lot of different items contribute to the Prosper score. So how good is
it in confirming to investors the estimated return?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot7}
ggplot(aes(x = ProsperScore, y = EstimatedReturn), 
       data = subset(pl, ProsperScore != "NA")) +
  geom_boxplot() +
# Focus plot view on most relevant values
  scale_y_continuous(breaks = seq(0, 0.30, 0.01), limits = c(0, 0.3))

summary(lm(EstimatedReturn ~ ProsperScore, data=pl))
```

Surprisingly the estimated return trends downward as prosper score increases 
past 6. Maybe this is because higher scores have higher loan amounts on average?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot8}
ggplot(aes(x = ProsperScore, y = LoanOriginalAmount),
       data = subset(pl, ProsperScore != "NA")) +
         geom_boxplot()

summary(lm(LoanOriginalAmount ~ ProsperScore, data=pl))
```

We do see a rough trend that direction, so while high prosper scores may have 
lower estimated returns the overall return could be greater due to the larger 
amount of money. Finally, for a bit of fun, let's take a look at what the loan 
amount is by state.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=8, Bivariate_Plot9}
qplot(x = LoanOriginalAmount, data = pl, binwidth = 2500) +
  scale_x_continuous(breaks = seq(0, 40000, 10000), limits = c(0, 36000)) +
  facet_wrap(~BorrowerState, ncol = 6) +
  x45
```

We can see that for the most part the individual states follow the same trend 
we saw in the overall graph earlier, though there are significantly more from 
California.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
There is a strong correlation between MonthlyLoanPayment and LoanOriginalAmount 
with MonthlyLoanPayment increasing as LoanOriginalAmount increases.

There are clear ceilings of LoanOriginalAmount in relation to AvergeCreditScore 
where a higher AverageCreditScore allows access to higher LoanOriginalAmounts.

There is a similar pattern between MonthlyLoanPayment and AverageCreditScore 
most likely due to higher LoanOriginalAmounts resulting in higher 
MonthlyLoanPayments.

There is some relationship between ProsperScore and AverageCreditScore, but the 
trend drops at the top ProsperScore of 11 which has a lower mean than a score 
of 10.

The majority of borrowers have low to 0 DelinquenciesLast7Years, but by looking 
at the chart there is a clear trend in declining delinquencies with higher 
ProsperScores.

The relationship between ProsperScore and DebtToIncomeRatio is similar to the 
relationship seen between ProsperScore and AverageCreditScore. Overall the 
DebtToIncomeRatio decreases as ProsperScore increases, but it goes back up at 
a ProsperScore of 11.

Surprisingly the relationship between DebtToIncomeRatio and 
DelinquenciesLast7Years is not what I expected. The number of 
DelinquenciesLast7Years increases going from a DebtToIncomeRatio of 0 to 0.15, 
but then it goes down and only slightly increases as the DebtToIncomeRatio 
approaches 1.

The EstimatedReturn decreases as the ProsperScore increases which is possibly 
due to them being lower risk investments

Similar to AverageCreditScore we see a general increase in OriginalLoanAmount 
as the ProsperScore increases.

There did not appear to be any relationship between BorrowerState and 
OrigilLoanAmount as each state appears to follow the same basic pattern as the 
overall data. There is a significantly larger amount of loans made to borrowers 
in California, but those numbers follow the same pattern.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
The most interesting relationship to me was the one between DebtToIncomeRatio 
and DelinquenciesLast7Years. I expected it to be a simple increasing trend as 
a higher ratio would make repaying debts more difficult and thus more 
delinquencies, but it appears that after a certain point those that maintain a 
high ratio also manage their finances well and so do not fall behind on debt 
payments.

### What was the strongest relationship you found?
The strongest relationship I found was the relationship between 
MonthlyLoanPayment and LoanOriginalAmount with an adjusted r-squared value of 
0.8686. This makes sense as the more money you borrow the higher the payments, 
and it shows that there is still some influence from other factors like 
AverageCreditScore and ProsperScore.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = ProsperScore), data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_brewer(type = 'seq', palette = 'RdYlBu')

ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = AverageCreditScore), 
       data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_gradient(low = 'red', high = 'blue', 
                       breaks = seq(400, 900, 100), limits = c(400, 900))

ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = LoanOriginationYear), data = subset(pl, 
                                                       ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_brewer(type = 'seq', palette = 'Blues')

ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = BorrowerAPR), data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_gradient(low = 'blue', high = 'red')

ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = EstimatedReturn), data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_gradient(low = 'red', high = 'blue', breaks = seq(0, 0.30, 0.05), 
                       limits = c(0, 0.3))

ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = TotalProsperLoans), data = subset(pl, 
                                                     ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_gradient(low = 'red', high = 'blue')

ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = Term), data = subset(pl, 
                                                     ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_brewer(type = 'seq', palette = 'RdYlGn')
```

I really wanted to try and determine what could be the source of the three 
distinct groupings when comparing monthly loan payment to the loan original 
amount. I worked my way through the variables I though to most likely be 
contributing to this pattern, going through Prosper score, average credit score 
loan origination year, APR, estimated return, total Prosper loans, and finally 
finding that the length of term for the loan matched the pattern with each 
section corresponding to one of the three term lengths (12, 36, and 60). There 
were some additional strong relationships shown with the estimated return 
APR but both were within the three sections established by the length of term. 
So with this information we can do a new facet wrapped set of plots to get a 
better view of how Prosper score matches the range of loan amount to monthly 
payment.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, Multivariate_Plots1}
ggplot(aes(x = MonthlyLoanPayment, y = LoanOriginalAmount, 
           color = ProsperScore), data = subset(pl, ProsperScore != "NA")) +
  geom_point() +
  scale_color_brewer(type = 'seq', palette = 'RdYlBu') +
  facet_wrap(~Term, ncol = 1)
```

This leads to data that looks like an excellent fit for a linear model, so I 
went through that process.

```{r echo=FALSE, message=FALSE, warning=FALSE, LinearModel}
lm1 <- lm(I(LoanOriginalAmount) ~ I(MonthlyLoanPayment),
          data = subset(pl, ProsperScore != "NA"))
lm2 <- update(lm1, ~ . + Term)
lm3 <- update(lm2, ~ . + ProsperScore)
mtable(lm1, lm2, lm3)
```

Another exploration I wanted to delve further into was the relation between 
the Prosper score of the borrower and the amount of money estimated as a return 
for the lender. Previously I had noted that the estimated return was lower for 
those with higher Prosper scores, and theorized that this would be offset by 
a higher loan amount and thus more money returned.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2}
ggplot(aes(x = ProsperScore, y = LoanOriginalAmount * EstimatedReturn),
       data = subset(pl, ProsperScore != "NA")) +
         geom_boxplot() +
# Limit scale to exclude extreem outliers
  scale_y_continuous(breaks = seq(-2000, 4000, 500), limits = c(-2000, 4000))
```

The differences between estimated and return and loan amount combine to make a 
more balanced range of the estimated amount of money an investor should expect 
back. There is still a bit of a trend that can be seen for mid range Prosper 
socores where it looks like the highest return amount is usually seen when 
investing on loans with a Prosper Score between 4 to 8 or 11. With this I would 
expect to see a pretty solid relationship between amount invested and estimated 
return for each individual investor.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3}
ggplot(aes(x = (LoanOriginalAmount/Investors), 
           y = ((LoanOriginalAmount * EstimatedReturn)/Investors), 
           color = Term),
           data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .3) +
  scale_color_brewer(type = 'seq', palette = 'RdYlGn')

ggplot(aes(x = (LoanOriginalAmount/Investors), 
           y = ((LoanOriginalAmount * EstimatedReturn)/Investors), 
           color = ProsperScore),
           data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .3) +
  scale_color_brewer(type = 'seq', palette = 'RdYlBu')

ggplot(aes(x = (LoanOriginalAmount/Investors), 
           y = ((LoanOriginalAmount * EstimatedReturn)/Investors), 
           color = Term),
           data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .5) +
  scale_color_brewer(type = 'seq', palette = 'RdYlGn') +
  facet_wrap(~ProsperScore, ncol = 3)
```

The initial plot showed a pretty strong relationship, but there is also a 
pattern that could indicate another influencing factor. After comparing against 
both term and Prosper score the pattern seems to be pretty clearly due to 
Prosper score with smaller influence from term. Again this final data looks like
a good fit for a linear model.

```{r echo=FALSE, message=FALSE, warning=FALSE, LinearModel2}
lm4 <- lm(I((LoanOriginalAmount * EstimatedReturn)/Investors) ~ 
            I(LoanOriginalAmount/Investors), 
          data = subset(pl, ProsperScore != "NA"))
lm5 <- update(lm4, ~ . + Term)
lm6 <- update(lm5, ~ . + ProsperScore)
mtable(lm4, lm5, lm6)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
While there was already a pretty strong relationship seen between 
LoanOriginalAmount and MonthlyLoanPayment in our previous analysis by using a 
facet wrap on Term and displaying colors by ProsperScore we get a cleaner and 
more structured result.

By combining EstimatedReturn and LoanOriginalAmount we can calculate the value 
in dollars of the EstimatedReturn. We see two distinct groups of values with 
ProsperScores 1-3 having low values due to a significant amount of loans with 
negative EstimatedReturn, and 4-11 having roughly even values. The pattern 
becomes more significant after including the value of the OriginalLoanAmount, 
not just the percentage return of it, showing an increasing trend that explains 
the difference between the EstimatedReturn and the calculated value of the 
return.

By further dividing the previous data by Investors we can find what an 
individual investor could expect to receive back depending on how much they 
invest. By further refining with ProsperScore and Term we see a very tight 
relationship outside of a few outliers.

### Were there any interesting or surprising interactions between features?
Of all the variables that influence the shape of the relationship between 
LoanOriginalAmount and MonthlyLoanPayment I was surprised to find that Term had 
the largest impact on the pattern. In hindsight it makes sense due to the 
limited range of Terms offered, as my assumption had been that they was much 
more variation in the values.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
Yes I created two models: one for the relationship between LoanOriginalAmount 
and MonthlyLoanPayment, and the other for the relationship between the per 
Investor return against the original cost.

For the model on LoanOriginalAmount and MonthlyLoanPayment it greatly benefited 
from the addition of Term, going from a r-squared value of 0.840 to 0.927. The 
inclusion of ProsperScore increased the r-squared value only slightly.

For the model between per Investor return (LoanOriginalAmount * 
EstimatedReturn)/Investors) and loaned amount (LoanOriginalAmount/Investors) the
initial comparison was strong with a r-squared value of 0.952. Adding 
ProsperScore and Term mildly increased the r-squared value.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = ProsperScore, y = AverageCreditScore, fill = ProsperScore),
       data = subset(pl, ProsperScore != "NA")) +
  scale_color_brewer(aesthetics = "fill", type = 'div', 
                     guide = guide_legend(title = 'Prosper Score'), 
                     palette = 'RdYlGn') +
  scale_y_continuous(breaks = seq(600, 900, 20)) +
  geom_boxplot() +
  labs(title = 'Average Credit Score by Prosper Score', 
       x = 'Prosper Score', y = 'Average Credit Score')
```

### Description One
This plot gives a good summary of where a credit score could land a borrower on 
the Prosper score scale with higher credit scores generally leading to a higher 
Prosper score, but there are some additional factors that come into play as seen
 by the lower but tighter grouping of credit scores for those with a Prosper 
score of 11 which is the highest.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, Plot_Two}
ggplot(aes(x = LoanOriginalAmount, y = MonthlyLoanPayment, 
           color = ProsperScore), 
       data = subset(subset(pl, ProsperScore != 'NA'),
                     MonthlyLoanPayment != 0)) +
  scale_x_continuous(breaks = seq(0, 36000, 2000)) +
  geom_point(alpha = .3) +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Prosper Score', 
                                    override.aes = list(alpha = 1, size = 2)),
                     palette = 'RdYlBu') +
  facet_wrap(~Term, ncol = 1) +
  labs(title = 'Monthly Payment by Loan Amount Faceted by Term in Months', 
       x = 'Loan Amount in Dollars', y = 'Monthly Payment in Dollars')
```

### Description Two
This plot shows the relationship between the monthly payment amount and the 
original borrowed amount in dollars along with the influence caused by the term 
of the loan along with the borrowers Prosper score. A long term and high Prosper
score result in the lowest payment amounts for a loan, and certain loan amounts 
are not available depending on the term of the loan.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, Plot_Three}
ggplot(aes(x = (LoanOriginalAmount/Investors), 
           y = ((LoanOriginalAmount * EstimatedReturn)/Investors), 
           color = ProsperScore),
           data = subset(pl, ProsperScore != "NA")) +
  geom_point(alpha = .3) +
  scale_x_continuous(breaks = seq(0, 36000, 2000)) +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Prosper Score', 
                                    override.aes = list(alpha = 1, size = 2)),
                     palette = 'RdYlGn') +
  labs(title = 'Estimated Individual Return by Individual Investment', 
       x = 'Initial Investment in Dollars',
       y = 'Estimated Individual Return in Dollars')
```

### Description Three
This plot shows the relationship between the calculated values for estimated 
individual return and the initial investment in dollars along with the influence
of the borrower's Prosper score on the result. This shows that safer 
investments, those with high Prosper scores, have a lower return rate and 
require a higher initial investment to have the same estimated return. In 
addition the highest estimated individual returns can only be seen with loans to
those with high Prosper scores.

------

# Reflection
With over 100 thousand records and 83 variables there was plenty of data to 
explore in the Prosper Loan set. After getting to know the variables I decided 
to approach the exploration from both the view of a borrower and a investor. I 
determined each view would have their own points of interest like determining 
the likely monthly loan payments for borrowers or analysis to suggest good 
investments for investors.

While exploring the data there were a couple difficulties I cam across. First 
was the large amount of variables which made it difficult at times to determine 
which variables would be worthwhile or relevant to explore. A good example of 
this was in my exploration of the relationship between loan amount and monthly 
payment amount. I could see there was another factor involved based on the plot 
but it took me a good amount of work to find the variable.

My second struggle was due to the range of the data. Part way through I found 
that there was a distinct split in the data around July 2009. After some 
exploration I found that most of the data was focused after this date, and so 
I filtered a lot of my explorations to exclude null values caused by this split 
and focus on those values after July 2009.

I feel my best success was in exploring the relationship between initial loan 
amount and monthly loan payment. By determining that the grouping I first 
noticed was caused by the term of the loan I was able to create a linear model 
for the data with an r-squared value of 0.939 which was a significant 
improvement of the value when just comparing the two base variables.

For future analysis I think more analysis and data could be gathered to better 
model not only expected returns but include actual returns for the loans. This 
could in turn be used to provide analysis of borrowers risk and return to 
possibly allow investors to better understand the likely result of lending to 
certain borrowers.